---
title: Classroom infection likelihood
author: Dan Kelley
date: "`r Sys.Date()`"
---

```{r echo=FALSE}
Pp <- 0.0345                           # BC 2021-08-12 positive/test
Pa <- 3625 / 5020302                   # BC 2021-08-12 cases/pop
n <- 10
Pi <- function(P, n)
{
    rval <- 100*(1-(1-P)^n)
    if (rval < 10) round(rval) else round(rval)
}
```

# Introduction

In recent weeks, major universities in both Canada [1] and the United States of
America [2] have policies mandating that students, faculty and staff have
COVID-19 vaccinations before entering campus.

Anyone with access to the news knows that many prominent business are similarly
enacting vaccine requirements, and the reason is a mystery to no-one (see
Figure 1, for example).  As was reported today, the chief medical officer of
Canada judges the nation to be in the midst of a fourth wave, and notes that
mandatory vaccinations are "on the table" [3].

Why is Dalhousie lacking behind?  In the virtual Q&A sessions on the topic [4],
the reason for not requiring vaccination was patient privacy.  But, surely,
this ought to be weighed against the risk of subjecting the people of Nova
Scotia to a new wave of COVID-19 infections, after so many months of sacrifice.

Or perhaps the idea is that the low infection rates in Nova Scotia imply a
lower risk here than elsewhere.  This is an argument that is difficult to make,
given recent infection trends through Canada and the rest of the world.  Nova
Scotia has low infection rates -- at this moment -- because of lockdown.

The consequence of opening up to thousands of outsiders seems to be a recipe
for disaster, based not just on tends seen elsewhere, but also based on the
wave of infections we experienced a few months ago, evidently sparked by just a
few individuals failing to practice quarantine.

Although there are many subtle issues involved in this problem, the basic
elements are straightforward enough for middle-school mathematics, as is
explained in the next section.

# A simple theory of transmission in a classroom setting

Let $n$ be the number of persons that a student encounters during a class.
With good ventilation, this will be a fraction of the class, perhaps just the
students in nearby seats, plus a few others who were close during entry to the
class.  A reasonable lower limit might be $n=10$, for the 8 students within one
seat, plus 2 more who were close during class entry^[Note that the $n$ values
are for class encounters only; there is also risk of exposure at home, etc.,
increasing the probabilities inferred here.]

The probability of not being exposed when two individuals meet is $1-P$, where
$P$ is the probability that a given person is infectious. Assuming independence
of encounter events, the probability of not being exposed if $n$ people
interact is $(1-P)^n$, and thus the probability of being exposed is
$1-(1-P)^n$.  This is the key formula for simple calculation of the probability
of being infected during an encounter.

Determining $P$ is difficult in the absence of extensive testing, knowledge of
whether infected individuals will attend classes, the time interval of
infectiousness, and other factors.  However, crude bounding values might be
provided by $P_a$, the ratio of active cases to population, and $P_p$, the
ratio of positive tests to total tests^[In the examples, $P_a$ and $P_p$
values are inferred from reported data for British Columbia on 2021-08-12.
These numbers appeared to be on the rise throughout North America, on this
date. Using present-day numbers for Nova Scotia is problematic, because so many
students come from out of province.].

**Example 1.** $n=`r n`$, $P_a=`r round(Pa,5)`$ and $P_p=`r Pp`$ suggests
that the probability of infection *in a given day* lies between `r Pi(Pa,n)`
and `r Pi(Pp,n)` percent.

```{r echo=FALSE}
n <- 40
```

**Example 2.** As example 1, but with $n=40$ to represent four classes (with
different neighbors), yields probabilities of being infected in a given day
from `r Pi(Pa,n)` to `r Pi(Pp,n)`%.

The second example is most relevant to the situation of a University student
taking a reasonable course load.

# Discussion

The calculations of the previous section suggest serious danger of infection,
using numbers appropriate in the middle of August, 2021. Infection rates being
on the rise lately in Canada (and much worse so in the USA), in-person teaching
seems likely to ignite a firestorm not just within Dalhousie, but in Nova
Scotia generally.

Of course, the model used here is very crude.  The province must have available
much more sophisticated models. But, do they take into account the $\Delta$
variant?  And do they even apply to a university environment, in which about a
quarter of the individuals are not from Canada, meaning that they tend to
arrive from places with higher infection rates.

The Dalhousie administrators who are tasked with making policy decisions about
going to in-class teaching ought to be asked for the results of their more
detailed and sophisticated models of this situation. If, as seems to be the
case, they are simply reusing calculations done by the province for encounters
in retail establishments, then there is a danger of poor prediction.  After
all, people do not shop for many hours each day, and they are not separated by
sub-metre distances when they do so.  And most of those shoppers are not as
socially active as university students tend to be.

If in-person teaching is to be the decision for Autumn 2021, what am I to tell
my students who ask me why vaccination is not required?  If I say it's to
respect privacy, they will ask why that does not apply to other major
universities and institutions.  And if I say it's because the chance of
infection is low in Nova Scotia, then either (a) they will ask me to explain
that mathematically, which I cannot do based on the previous section or (b)
worse, and a real possibility early in the term, they may point out that Nova
Scotia is already in a COVID-19 surge.

Dalhousie takes pride in its contributions to the community. Will it be proud
to be the cause of suffering in the province that has supported it for two
centuries?


![New daily cases in Canada, as of 2021-08-11. The most recent daily rate is written after the date, and black dots are used for the 10 most recent observations. The magenta line results from the use of a smoothing cubic spline, and is included to guide the eye through the "noisy" data. This figure was created based on data downloaded from <https://health-infobase.canada.ca/src/data/covidLive/covid19.csv> on 2021-08-12, and shows data up to the previous day.](covid_20210812.png)

# References

1. News Staff. “Western University, U of T to Mandate COVID-19 Vaccines for All
   Students, Staff - CityNews Toronto,” August 11, 2021.
   https://toronto.citynews.ca/2021/08/11/western-university-covid-19-vaccine-students-staff/.

2. University of California (Cnet). “UC Issues Final COVID-19 Vaccination
   Policy | UCnet,” July 15, 2021.
   https://ucnet.universityofcalifornia.edu/news/2021/07/ucs-covid-19-vaccine-policy.html.

3. Tasker, John Paul. “As Delta Variant Spreads, Canada’s Top Doctors Say
   Mandatory Vaccination for Some Workers Is on the Table | CBC News.” CBC,
   August 12, 2021.
   https://www.cbc.ca/news/politics/delta-variant-spreads-mandatory-vaccination-1.6139102.

4. Dalhousie University. “Virtual Q&A Discussion Videos,” July 20, 2021.
   https://www.dal.ca/covid-19-information-and-updates/our-approach/virtual-q-a-discussion-videos.html.
