---
title: Dalhousie Vaccine Policy in the Context of Classroom Infection
author: Dan Kelley, Professor, Department of Oceanography (dan.kelley@dal.ca)
date: "`r Sys.Date()`"
header-includes:
   - \usepackage{graphicx}
   - \usepackage{marginnote}
   - \usepackage{times}
   - \usepackage{amsmath}
   - \usepackage{microtype}
---

```{r echo=FALSE}
Pp <- 0.0345                           # BC 2021-08-12 positive/test
Pa <- 3625 / 5020302                   # BC 2021-08-12 cases/pop
n <- 10
Pi <- function(P, n)
{
    rval <- 100*(1-(1-P)^n)
    paste0(if (rval < 10) round(rval) else round(rval), "%")
}
```

# Introduction

In recent weeks, major universities in both Canada [1] and the United States of
America [2] have enacted policies requiring that students, faculty and staff
have COVID-19 vaccinations before entering campus.

Anyone with access to the news knows that many prominent businesses are also
enacting vaccine requirements. The reason is a mystery to no-one.  As was
reported yesterday, the chief medical officer of Canada judges the nation to be
in the midst of a fourth wave, and notes that mandatory vaccinations are "on
the table" [3]. For any readers that might need a reminder, Figure 1 on the
next page paints an sobering picture.


Why is Dalhousie lagging behind?  In the virtual Q&A sessions on the topic [4],
the reason stated for not requiring vaccination was patient privacy.  But,
surely, privacy must be weighed against the risk of subjecting Nova Scotians to
unnecessary misery, after so many months of sacrifice.

Or perhaps the idea is that the low infection rates in Nova Scotia imply a
lower risk here than elsewhere.  This is a flimsy argument, given recent
infection trends through Canada and the rest of the world.

Nova Scotia has low infection rates at this particular moment, but we all know
why: travel restriction.  It is not a coincidence that so many cases are
travel-related.

The consequence of opening up to thousands of outsiders seems to be a recipe
for disaster, based on trends seen elsewhere, and also on the frightening wave
that occurred a few months ago, which may have been sparked by just a few
individuals failing to practice quarantine.

The university classroom is a special place, not quite the same as a mall or a
museum.  Although there are many subtle issues involved in considering COVID-19
spread in a lecture hall, the basic elements are simple enough to be explained
with middle-school mathematics, as in the next section.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=\hsize]{covid_letter_figure.pdf}
\end{center}
\caption{\footnotesize New daily cases in Canada, as of 2021-08-12. The most recent daily rate is written after the date, and black dots indicate the 10 most recent observations. The magenta lines, shown to guide the eye through "noisy" data, are computed using a smoothing cubic spline. This figure is based on data downloaded from <https://health-infobase.canada.ca/src/data/covidLive/covid19.csv> on 2021-08-13, and shows observations ranging up to the previous day.}
\end{figure}


# A simple model of classroom disease transmission

Let $n$ be the number of persons that a student encounters during a class.
With good ventilation, this will not equal the total number of students in the
class, so we can set aside the notion that $n$ is 400 or more in some classes.
It is more reasonable to take $n$ to be the number of students in nearby seats,
plus a few others who were close during class entry and exit.  A reasonable
lower limit might be $n=10$, for the 8 students in neighbouring seats, plus 2
more who were close at the door or in the hallway^[The $n$ values are for class
encounters only; there is also risk of exposure at home, etc., and this will
increase the probabilities inferred here.].

Let $P$ be the probability that a given person is infectious, and consider the
contact between two individuals, one of whom is not infected, the other
possibly being infectious. Supposing sufficient exchange of virus load^[The
rate of exchange involves disease properties, air flow, the extent of cleaning
between classes, etc., and is beyond the present scope.], the probability that
the uninfected individual becomes infected is $1-P$ per encounter.  Assuming
independence of encounter events, the probability of not being exposed if $n$
people interact is given $(1-P)^n$.  It follows that the probability of being
infected during an $n$-person interaction is $1-(1-P)^n$. This simple
formula^[If $P\ll 1$ and $n$ is small, the formula may be approximated as $nP$,
which permits calculation by eye.] is the key to inferring infection
probability during a lecture.

Determining $P$ is difficult in the absence of extensive testing, knowledge of
whether infected individuals will attend classes, the time interval of
infectiousness, and other factors.  However, crude bounding values might be
provided by $P_a$, the ratio of active cases to population, and $P_p$, the
ratio of positive tests to total tests. In the examples below, $P_a$ and $P_p$
are inferred from reported data for British Columbia on 2021-08-12, which may
provide plausible estimates of the Nova Scotian situation a few weeks into
classes, if past is prologue^[A refinement to the present calculation might
break down Dalhousie students by point of origin, but this is not information
available to the author.].

**Example 1.** $n=`r n`$, $P_a=`r round(Pa,5)`$ and $P_p=`r Pp`$ suggests an
infection probability between `r Pi(Pa,n)` and `r Pi(Pp,n)`.

```{r echo=FALSE}
n <- 40
```

**Example 2.** A similar calculation, but with $n=40$ to represent four classes
(with different neighbors), yields infection probabilities between `r Pi(Pa,n)`
and `r Pi(Pp,n)`.

The first example might apply to a part-time student who takes just one class,
while the second is more relevant to a full-time student.

Whether even the lower estimates are acceptable is an open question, given that
these are computations just for a single encounter, and probabilities would be
higher if multiple encounters are assumed, with differing classroom
neighbors^[In my own teaching, I have noticed that it takes a week or two for
students to settle on a spot where they will habitually sit.].  Extending this
model to such situations would not be difficult, and I do hope that Dalhousie's
administrators have undertaken such calculations.

# Discussion

Dalhousie plans to return to in-person teaching in a few weeks, if Nova Scotia
enters phase 5, and I certainly look forward to seeing my students in person
again. My guess is that they will all be wearing masks, required or not. But
what am I to say if they why vaccination is not required? I seem to have three
choices.

* If I say it's to respect privacy, they may ask me why this moral imperative
  does not apply to other universities, institutions, and even nations.  I have
  no answer to that.

* If I say it's because the chance of infection is low in Nova Scotia, then
  I fear my students might respond in either of two ways. They might ask me to
  explain that mathematically, which is difficult because I think the reverse
  is true.  Or they might simply point to the current case load, which we all
  hope will be low, but fear may be high.

* If I say it's because of technical difficulties, they will respond that
  Canada has set up a system for vaccination passes [5], so very little effort
  would be required on the part of Dalhousie.

My honest response will be that I do not understand the Administration's
choice, and cannot defend it.

I grew up in Nova Scotia, and I know that Dalhousie has long taken pride in its
contributions to the community. I wonder, will it be proud to be the cause of
unnecessary suffering in a province that has supported it for two centuries?

Demanding that students, staff and faculty be vaccinated is not just a moral
duty; it is a practical one, for an institution that relies on community
support to survive.



# References

1. News Staff. “Western University, U of T to Mandate COVID-19 Vaccines for All
   Students, Staff - CityNews Toronto,” August 11, 2021.
   https://toronto.citynews.ca/2021/08/11/western-university-covid-19-vaccine-students-staff/.

2. University of California (Cnet). “UC Issues Final COVID-19 Vaccination
   Policy | UCnet,” July 15, 2021.
   https://ucnet.universityofcalifornia.edu/news/2021/07/ucs-covid-19-vaccine-policy.html.

3. Tasker, John Paul. “As Delta Variant Spreads, Canada’s Top Doctors Say
   Mandatory Vaccination for Some Workers Is on the Table | CBC News.” CBC,
   August 12, 2021.
   https://www.cbc.ca/news/politics/delta-variant-spreads-mandatory-vaccination-1.6139102.

4. Dalhousie University. “Virtual Q&A Discussion Videos,” July 20, 2021.
   https://www.dal.ca/covid-19-information-and-updates/our-approach/virtual-q-a-discussion-videos.html.

5. Tunney, Catharine. “Ottawa Promises Vaccine Passport for International
   Travel This Fall | CBC News.” CBC, August 11, 2021.
   https://www.cbc.ca/news/politics/vaccine-passport-1.6137763.

