---
title: Dalhousie Vaccine Policy in the Context of Classroom Infection
author: Dan Kelley, Professor, Department of Oceanography (dan.kelley@dal.ca)
date: "`r Sys.Date()`"
---

```{r echo=FALSE}
Pp <- 0.0345                           # BC 2021-08-12 positive/test
Pa <- 3625 / 5020302                   # BC 2021-08-12 cases/pop
n <- 10
Pi <- function(P, n)
{
    rval <- 100*(1-(1-P)^n)
    if (rval < 10) round(rval) else round(rval)
}
```

# Introduction

In recent weeks, major universities in both Canada [1] and the United States of
America [2] have enacted policies requiring that students, faculty and staff
have COVID-19 vaccinations before entering campus.

Anyone with access to the news knows that many prominent business are similarly
enacting vaccine requirements, and the reason, clear evidence of a fourth wave,
should be a mystery to no-one.  As was reported today, the chief medical
officer of Canada judges the nation to be in the midst of a fourth wave, and
notes that mandatory vaccinations are "on the table" [3]; Figure 1 shows data
broken down by province.

Why is Dalhousie lacking behind?  In the virtual Q&A sessions on the topic [4],
the reason for not requiring vaccination was patient privacy.  But, surely,
this ought to be weighed against the risk of subjecting the people of Nova
Scotia to a new wave of COVID-19 infections, after so many months of sacrifice.

Or perhaps the idea is that the low infection rates in Nova Scotia imply a
lower risk here than elsewhere.  This is a flimsy argument, given recent
infection trends through Canada and the rest of the world.  Nova Scotia has low
infection rates at this moment, but anyone living here knows why: lock-down.

The consequence of opening up to thousands of outsiders seems to be a recipe
for disaster, based not just on tends seen elsewhere, but also based on the
wave of infections we experienced a few months ago, perhaps sparked by just a
few individuals failing to practice quarantine.

Although there are many subtle issues involved in this problem, the basic
elements are straightforward enough to be explained using middle-school
mathematics, as shown in the next section.

# A simple theory of transmission in a classroom setting

Let $n$ be the number of persons that a student encounters during a class.
With good ventilation, this will be a fraction of the class, perhaps just the
students in nearby seats, plus a few others who were close during entry to the
class.  A reasonable lower limit might be $n=10$, for the 8 students within one
seat, plus 2 more who were close during class entry^[Note that the $n$ values
are for class encounters only; there is also risk of exposure at home, etc.,
and this will increase the probabilities inferred here.]

Let $P$ be the probability that a given person is infections, and consider the
contact between two individuals, one of whom is not infected. Supposing
sufficient exchange of virus load the probability that the uninfected
individual pickup up infection is $1-P$ per encounter. Assuming independence of
encounter events, the probability of not being exposed if $n$ people interact
is given $(1-P)^n$.

It follows that the probability of being infected during an $n$-person
interaction is $1-(1-P)^n$. This simple formula is the key to inferring
infection probability during a lecture.

Determining $P$ is difficult in the absence of extensive testing, knowledge of
whether infected individuals will attend classes, the time interval of
infectiousness, and other factors.  However, crude bounding values might be
provided by $P_a$, the ratio of active cases to population, and $P_p$, the
ratio of positive tests to total tests. In the examples below, $P_a$ and $P_p$
are inferred from reported data for British Columbia on 2021-08-12, and there
is no reason to think they are underestimates of where Nova Scotia might be a
few weeks into classes. A refinement to the present calculation should break
down Dalhousie students by point of origin, but this is not information
available to the author.  There is also no firm way to guess what conditions
will be like anywhere in a month, although the British Columbia infection curve
may be a relevant predictor, given that it has a similar health system to that
in Nova Scotia.

**Example 1.** $n=`r n`$, $P_a=`r round(Pa,5)`$ and $P_p=`r Pp`$ suggests
that the probability of infection *in a given day* lies between `r Pi(Pa,n)`
and `r Pi(Pp,n)` percent.

```{r echo=FALSE}
n <- 40
```

**Example 2.** As example 1, but with $n=40$ to represent four classes (with
different neighbors), yields probabilities of being infected in a given day
from `r Pi(Pa,n)` to `r Pi(Pp,n)`%.

The second example is most relevant to the situation of a University student
taking a reasonable course load.

# Discussion

The calculations of the previous section suggest serious danger of infection,
using numbers appropriate in the middle of August, 2021. Infection rates being
on the rise lately in Canada (and much worse so in the USA), in-person teaching
seems likely to ignite a firestorm not just within Dalhousie, but in Nova
Scotia generally.

Of course, the model used here is very crude.  The province must have available
much more sophisticated models. It seems that Dalhousie is relying on models
calibrated to the Springtime conditions, but evidence around the world has
revealed that the parameters need adjustment for the "Delta" variant, and so a
Dalhousie decision made months ago might need to be revisited now.  Besides,
one might ask whether those models apply to a university environment, with its
close and repeated interactions between multiple groups.

If in-person teaching is to be the decision for Autumn 2021, what am I to tell
my students who ask me why vaccination is not required? There seem to be two
choices.

* If I say it's to respect privacy, they may ask me why this moral imperative
  does not apply to other universities, institutions, and even nations.  I have
  no answer to that.

* If I say it's because the chance of infection is low in Nova Scotia, then
  I fear my students might respond in either of two ways. If they ask me to
  explain that mathematically, I will fail, because, as explained above, I
  believe the reverse.

\bigskip
\bigskip
\bigskip

Dalhousie has long taken pride in its contributions to the community. Will it
be proud to be the cause of suffering in the province that has supported it for
two centuries?  Demanding that students, staff and faculty be vaccinated is not
just a moral duty, it is also a politically sensible decision.


![New daily cases in Canada, as of 2021-08-11. The most recent daily rate is written after the date, and black dots are used for the 10 most recent observations. The magenta line results from the use of a smoothing cubic spline, and is included to guide the eye through the "noisy" data. This figure was created based on data downloaded from <https://health-infobase.canada.ca/src/data/covidLive/covid19.csv> on 2021-08-12, and shows observations ranging up to the previous day.](covid_20210812.png)

# References

1. News Staff. “Western University, U of T to Mandate COVID-19 Vaccines for All
   Students, Staff - CityNews Toronto,” August 11, 2021.
   https://toronto.citynews.ca/2021/08/11/western-university-covid-19-vaccine-students-staff/.

2. University of California (Cnet). “UC Issues Final COVID-19 Vaccination
   Policy | UCnet,” July 15, 2021.
   https://ucnet.universityofcalifornia.edu/news/2021/07/ucs-covid-19-vaccine-policy.html.

3. Tasker, John Paul. “As Delta Variant Spreads, Canada’s Top Doctors Say
   Mandatory Vaccination for Some Workers Is on the Table | CBC News.” CBC,
   August 12, 2021.
   https://www.cbc.ca/news/politics/delta-variant-spreads-mandatory-vaccination-1.6139102.

4. Dalhousie University. “Virtual Q&A Discussion Videos,” July 20, 2021.
   https://www.dal.ca/covid-19-information-and-updates/our-approach/virtual-q-a-discussion-videos.html.
